---
title: "DoOR analysis tools"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The `DoOR.functions` package provides tools for the analysis of the data provided by the `DoOR.data` package.

## Content
* [Loading data](#loading)
* [Identifying the sensillum we are recording from with `identifySensillum()`](#identifySensillum)
* [Finding neuron-specific odorants with `privateOdorant()`](#privateOdorant)
* [Mapping response data from an unknown source with `mapReceptor()`](#mapReceptor)
* [Changing the response unit with `backProject()`](#backProject)
* [Quantifying tuning curves with `sparse()`](#sparse)



## Loading data{#loading}
First we need to load packages and data:
```{r, echo = T, message = TRUE, results='hide'}
#load data
library(DoOR.functions)
library(DoOR.data)
loadData()
```

## Identifying the sensillum we are recording from with `identifySensillum()` {#identifySensillum}
Imagine we perform an electrophysiological recording from a _Drosophila_ sensillum (single sensillum recording, SSR) and we are not sure what sensillum we are recording from. In order to identify the sensillum we used several diagnostic odorants (maybe selected using [`privateOdorant()`](#privateOdorant)) and got responses from the different sensory neurons the sensillum houses. We can now pass our recorded data to `identifySensillum()`.

Let's make up some simple fake data. We pretend to have recorded with three odorants (2,3-butanedione, ethanoic acid and carbon dioxide) and we could separate the responses of two units. Unit1 responded strongly to 2,3-butanedione only, unit2 only responded to carbon dioxide. We create a data.frame that contains a column called `odorants` with the InChIKeys of our test odorants, and one column for each unit (name the colnames as you like, e.g. unit1-n or Aneuro if you are sure about the neuron). 

```{r}
recording <- data.frame(
  odorants = c(transID(c("BEDN", "ETAS"), "Code"), 
               transID("carbon dioxide", "Name")), 
  unit1 = c(.9,.1,.1), 
  unit2 = c(0, .1, 1)
)
```
Next we feed the recording to the function:

### using correlations
```{r, fig.width=7.1, fig.height=4.5}
identifySensillum(recording, base_size = 8)
```
Note that the function tells us that it found hits for all units in ab1 and ab5, meaning that e.g. within the four neurons housed in the ab1 sensillum both of our units had good matches. You can set this correlation theshold with `min.cor`. If we increase the threshold to 0.99 only ab1 is returned as a double match:

```{r, fig.width=7.1, fig.height=4.5, fig.show='hide'}
identifySensillum(recording, min.cor = .99)
```

We can define the number of best hits that we want to get returned (the default is 10):
```{r, fig.width=7.1, fig.height=4.5}
identifySensillum(recording, nshow = 5, base_size = 8)
```
And if we know e.g. that we are recording from a basiconic sensillum we can restrict the search to one or a few sensillum types:
```{r, fig.width=7.1, fig.height=4.5}
identifySensillum(recording, sub = "ab", nshow = 5, base_size = 8)
identifySensillum(recording, sub = c("ac","at"), nshow = 5, base_size = 8)
```

### using Euclidean distances
Instead of correlations we can also use the Euclidean distance as a (dis)similarity measure:

```{r, fig.width=7.1, fig.height=4.5}
identifySensillum(recording, method = "dist", sub = "ab", nshow = 5, base_size = 8)
```

### returning data instead of plots
We can also return the correlation/distance data instead of the plot when setting `plot =FALSE`:
```{r}
sensillumX <- identifySensillum(recording, method = "dist", sub = "ab", plot = F)
head(sensillumX)
```

So apparently our fake recording came from the ab1 sensillum, which was admittedly quite obvious as we had a strong carbon dioxide response and ab1 houses the carbon dioxide receptor :)



## Finding neuron-specific odorants with `privateOdorant()` {#privateOdorant}
There may be several cases where we might be interested in so called _private odorants_, odorants that specifically activate a given receptor or sensory neuron. Maybe we are looking for diagnostic odorants for sensillum identification or we want to activate a specific neuronal pathway, `privateOdorant()` returns candidate odorants for that task.

Let's say we want to specifically activate Or22a neurons:
```{r}
privateOdorant("Or22a")
```

We might want to return the odorant names instead of InChiKeys:
```{r}
privateOdorant("Or22a", tag = "Name")
```
So according to the function sec-amyl acetate would be a good candidate. It activates Or22a at 0.4 (DoOR response, max is 1) while the maximum activation in all other tested responding units (receptors, neurons, glomeruli) is 0.016, a difference of 0.40. Sounds good, but it was tested only in 4 other responding units, so I would rather go for ethyl hexanoate with about the same difference but being tested in 29 other responding units.

We can also restrict the search to the sensillum the responding units of interest is related to:
```{r}
privateOdorant("Or22a", tag = "Name", sensillum = T)
```
Ethyl 2-methylbutanoate sounds like a good hit, it has the same difference to the other units as ethyl hexanoate but hardly elicits a response at all from the other neuron. The n of 1 is fine as there are only 2 neurons housed in the ab3 sensillum.



## Mapping response data from an unknown source with `mapReceptor()`{#mapReceptor}
Similar to `identifySensillum()`, `mapReceptor()` correlates a response vector to all responding units of the  existing DoOR consensus data. Let's grab a data set from Or22a and see where it ends up:
```{r}
data <- data.frame(odorants  = Or22a$InChIKey, responses = Or22a$Hallem.2006.EN)
data <- na.omit(data)
head(data)
mapReceptor(data = data, nshow = 5)
```
This example was a bit circular as the tested data contributed to the consensus data...



## Changing the response unit (spikes, deltaF/F, ...) with `backProject()`{#backProject}
The DoOR consensus data is normalized to values between 0 and 1. If we want to compare the DoOR data to one of our own recordings, it would be great to have the DoOR data in the same unit as our owne data (e.g. spikerate). This is what we can do with `backProject()`,  we can rescale the DoOR data to fit a given response template.

As an example, let's take the data Hallem et al. recorded from Or22a _via_ calcium imaging and rescale the DoOR responses accordingly. The template has to have 2 columns named `odorants` and `responses`:
```{r, fig.width = 4.5, fig.height = 4.5}
template <- data.frame(odorants  = Or22a$InChIKey, 
                       responses = Or22a$Hallem.2006.EN)

bp <- backProject(template, responding.unit = "Or22a")
plot(bp$backprojected$original.data, 
     bp$backprojected$backprojected.data, 
     xlab = "DoOR consensus response", 
     ylab = "backprojected data [spikes, Hallem.2006.EN]"
)

head(bp$backprojected)

```
All the yellow lines in the first plot represent odorantt responses that were not available in the original data set but were projected onto the fitted function and rescaled to the units in "Hallem.2006.EN". The second plot shows the relationship between the rescaled data and the original consensus responses.


## Quantifying tuning curves with `sparse()`{#sparse}
The width of a tuning curve, i.e. for examplt to how many odorants a receptor shows strong responses, can be quantified using different sparseness measures. We implemented kurtosis[^1] and sparseness[^2] in `sparse()`. A high kurtosis or sparseness value indicates a narrow tuning curve (see also `DoORplot_tuningCurve()` in the [DoOR visualization vignette](DoOR_visualizations.html#tuningCurve)). While kurtosis is able to deal with negative values, sparseness can't and thus all values will be transformed to absolute values. Sparseness scales between 0 and 1, kurtosis between -∞ and ∞, a kurtisis of 0 corresponds to the Gaussian distribution.

```{r}
rm.SFRreset <- resetSFR(response.matrix, "SFR")

sparse(x = rm.SFRreset[,"Or69a"], method = "ltk")
sparse(x = rm.SFRreset[,"Or69a"], method = "lts")

sparse(x = rm.SFRreset[,"Gr21a.Gr63a"], method = "ltk")
sparse(x = rm.SFRreset[,"Gr21a.Gr63a"], method = "lts")

```



[^1]: Willmore, B., Tolhurst, D.J., 2001. Characterizing the sparseness of neural codes. Network 12, 255–270. doi:10.1080/net.12.3.255.270
[^2]: Bhandawat, V., Olsen, S.R., Gouwens, N.W., Schlief, M.L., Wilson, R.I., 2007. Sensory processing in the Drosophila antennal lobe increases reliability and separability of ensemble odor representations. Nature neuroscience 10, 1474–82. doi:10.1038/nn1976
